### 2024-09-23 paper/paper.qmd
Edits to Paper.qmd

---
title: "Toronto's Covid-19 Impact: How did Age and Gender affect Outcomes?"
author: 
  - Dhruv Gupta
thanks: "Code and data are available at: https://github.com/dhruv5423/Covid19-R-Project"
date: "today"
date-format: long
abstract: "The Covid-19 Pandemic caused irreparable damage to the infrastructure of our global systems, highlighting the neccessity to better understand factors that influence health outcomes. This paper uses data from OpenDataToronto to investigate how age and gender affects outcomes of those infected with the virus. xyz about results"
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

#Loading Packages
library(tidyverse)
library(here)
library(palmerpenguins)
library(opendatatoronto)
library(tinytex)

#Loading Data
raw_covid_data <- read.csv(here("data/raw_data/raw_covid_data.csv"))
cleaned_covid_data <- read.csv(here("data/analysis_data/cleaned_covid_data.csv"))
```

# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it.

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

```{=tex}
\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}
```
We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {.unnumbered}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

\newpage

# References

### 2024-09-23 paper/paper.qmd
edits to paper.qmd and references

@@
 format: pdf
+toc: true
 number-sections: true
@@
 
-You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.
+The Covid-19 Pandemic has had an unimaginable effect on human lives around the globe.
+As of April 13, 2024, Worldometer Info estimates that over 704 million people worldwide have contracted the virus, resulting in approximately 7 million deaths [@worldometer].
+In Canada alone, there have been almost 4.6 million reported cases and more than 38,000 deaths as of July 20, 2024 [@infobase_ca].
 
-The remainder of this paper is structured as follows. @sec-data....
+While these statistics paint a harrowing picture of the human toll inflicted, the pandemic has exacerbated existing economic inequalities, destabilized political systems, and put immense pressure on societal infrastructure globally.
+Looking past the immediate health crisis, lockdowns and restrictions have had immeasurable impacts not only on global supply chains, but also on the mental heatlh of many forced to quarantine or self isolate.
+
+An article by Fortune valued the economic burden on the US Economy to be upwards of \$14 Trillion USD at the end of 2023 [@fortune].
+While this figure mainly took into account the 'standard economic effects' of the pandemic - revenue lost due to mandatory business closures, decreases in air travel, and workplace absences - the article noted that there were many unobservable factors that were incredibly burdernsome to the economy, such as long term physical and mental health effects of the pandemic on the population.
+
+Covid-19 has been found to have varying effects across demographics.
+A 2020 article published in the PLOS Journal found that 'Covid-19 may be associated with worse outcomes in males than in females'.
+The article found that men are up to 22% more likely to require ICU admission.
+
+Moreover, an article published in the Springer Link Journal in 2021 found that older adults, in particular those above the age of 65, face higher mortality rates than their younger counterparts.
+Weaker immune systems, and the higher likely presence of other conditions can exacerbate the effects of the virus.
+
+Understanding how demographic variables like age and gender affect outcomes related to contracting viruses is increasingly important in the shaping of future policies and health measures.
+This paper aims to analyse the differences in outcomes for various age groups and genders among Covid-19 cases in Toronto, in an effort to contribute to deepening our understanding of the risk factors that may impact the lives of those with Covid-19, and possibly in future pandemics as well.
+
+The remainder of this paper is structured as follows. @sec-data xyz @sec-results xyz @sec-discussion xyz.
 
@@
 
-Some of our data is of penguins (@fig-bills), from @palmerpenguins.
+We use @citeR and @rohan.
 
-```{r}
-#| label: fig-bills
-#| fig-cap: Bills of penguins
-#| echo: false
+## Data Selection and Measurement
 
-ggplot(penguins, aes(x = island, fill = species)) +
-  geom_bar(alpha = 0.8) +
-  scale_fill_manual(values = c("darkorange","purple","cyan4"),
-                    guide = "none") +
-  theme_minimal() +
-  facet_wrap(~species, ncol = 1) +
-  coord_flip()
-```
+## Raw Data
 
-Talk more about it.
-
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
-
-```{r}
-#| label: fig-planes
-#| fig-cap: Relationship between wing length and width
-#| echo: false
-#| warning: false
-#| message: false
-
-analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
-
-analysis_data |> 
-  ggplot(aes(x = width, y = length)) +
-  geom_point(alpha = 0.8) +
-  theme_minimal() +
-  labs(x = "Wing width (mm)",
-       y = "Wing length (mm)")
-```
+## Data Cleaning
 
@@
 
-# Model
+## Summary Statistics
 
-The goal of our modelling strategy is twofold. Firstly,...
+# Results {#sec-results}
 
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
+## Cases over Time
 
-## Model set-up
+## Age vs Outcomes
 
-Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
+## Gender vs Outcomes
 
-```{=tex}
-\begin{align} 
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
-\end{align}
-```
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
+# Discussion {#sec-discussion}
 
-### Model justification
-
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
-
-We can use maths by including latex between dollar signs, for instance $\theta$.
-
-# Results
-
-Our results are summarized in @tbl-modelresults.
-
-```{r}
-#| echo: false
-#| eval: true
-#| warning: false
-#| message: false
-
-library(rstanarm)
-
-first_model <-
-  readRDS(file = here::here("models/first_model.rds"))
-```
-
-```{r}
-#| echo: false
-#| eval: true
-#| label: tbl-modelresults
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
-#| warning: false
-
-modelsummary::modelsummary(
-  list(
-    "First model" = first_model
-  ),
-  statistic = "mad",
-  fmt = 2
-)
-```
-
-# Discussion
-
-## First discussion point {#sec-first-point}
+## Effect of Age on Outcomes
 
@@
 
-## Second discussion point
+## Effect of Gender on Outcomes
 
-## Third discussion point
+## Broader Discussion and Takeaways
 
@@
 
-\appendix
+# References
 
-# Appendix {.unnumbered}
-
-# Additional data details
-
-# Model details {#sec-model-details}
-
-## Posterior predictive check
-
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
-
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
-
-```{r}
-#| eval: true
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-ppcheckandposteriorvsprior
-#| layout-ncol: 2
-#| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
-
-pp_check(first_model) +
-  theme_classic() +
-  theme(legend.position = "bottom")
-
-posterior_vs_prior(first_model) +
-  theme_minimal() +
-  scale_color_brewer(palette = "Set1") +
-  theme(legend.position = "bottom") +
-  coord_flip()
-```
-
-## Diagnostics
-
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
-
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
-
-```{r}
-#| echo: false
-#| eval: true
-#| message: false
-#| warning: false
-#| label: fig-stanareyouokay
-#| fig-cap: "Checking the convergence of the MCMC algorithm"
-#| fig-subcap: ["Trace plot", "Rhat plot"]
-#| layout-ncol: 2
-
-plot(first_model, "trace")
-
-plot(first_model, "rhat")
-```
-
-\newpage
-
-# References
+# LLM Disclosure

### 2024-09-24 paper/paper.qmd
edits to cleaned data, cleaning script, testing script, and paper

@@
 ---
-title: "Toronto's Covid-19 Impact: How did Age and Gender affect Outcomes?"
+title: "Toronto's Covid-19 Impact: Effect of Age and Gender on Outcomes"
 author: 
@@
 bibliography: references.bib
+header-includes:
+  - \usepackage{float}
 ---
@@
 #Loading Packages
+library(pander)
 library(tidyverse)
@@
 library(tinytex)
+library(knitr)
+library(kableExtra)
 
@@
 ```
+
+\newpage
 
@@
 
-The remainder of this paper is structured as follows. @sec-data xyz @sec-results xyz @sec-discussion xyz.
+The remainder of this paper is structured as follows. @sec-data discusses @sec-results xyz @sec-discussion xyz.
+
+\newpage
 
@@
 
-## Data Selection and Measurement
+## Data Selection
+
+Data used in this report was sourced from OpenDataToronto's portal.
+@opendatatoronto.
+More specifically, the dataset "Covid-19 Cases in Toronto" was used and cleaned for the purposes of this report.
+Toronto Public Health (TPH) released anonymized, person-level data from the start of the pandemic in January 2020.
+The data spans from the first reported case on in January of 2020 to February 14th, 2024.
+In a statement on the website for this dataset, OpenDataToronto states that "As case and outbreak management guidelines changed and COVID-19 specific resources were no longer funded, the level of detail available for cases decreased, and more recent data are less complete and not comparable to previous years.
+TPH discontinued the production of this report with the final refresh as of February 14, 2024" @opendatatoronto_set
 
@@
 
+In it's original form, the dataset contains more than 414,000 entries regarding information on cases of Covid-19 in Toronto.
+Below are two tables with the first three rwos of the raw data, separated into two tables for readability.
+
+### Table 1: Sample of Raw Covid-19 Data
+
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+library(kableExtra)
+
+# First part of Table 1 (first few columns)
+raw_covid_data %>%
+  select(X_id, Outbreak.Associated, Assigned_ID, Age.Group, Neighbourhood.Name, FSA, Source.of.Infection, Classification) %>%
+  head(3) %>%
+  knitr::kable(booktabs = TRUE) %>%
+  kable_styling(font_size = 8)
+
+# Second part of Table 1 (next set of columns)
+raw_covid_data %>%
+  select(Episode.Date, Reported.Date, Client.Gender, Outcome, Ever.Hospitalized, Ever.in.ICU, Ever.Intubated) %>%
+  head(3) %>%
+  knitr::kable(booktabs = TRUE) %>%
+  kable_styling(font_size = 8)
+
+```
+
+(\scriptsize Table Separated using Tips from [@stack_tip] and LLMs)
+
 ## Data Cleaning
 
-Talk way more about it.
+### Table 2: Sample of Cleaned Covid-19 Data
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+# Cleaned Data Table (Table 2)
+cleaned_covid_data %>%
+  head(5) %>%
+  kable(col.names = c("x_id", "age_group", "client_gender", "reported_date", "ever_in_icu", "ever_hospitalized", "ever_intubated", "outcome"),
+        booktabs = TRUE)
+```
 
 ## Summary Statistics
+
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+cleaned_covid_data %>%
+  summarise(
+    total_cases = n(),
+    age_group_distribution = table(age_group),
+    gender_distribution = table(client_gender),
+    outcome_distribution = table(outcome),
+    earliest_reported_date = min(reported_date, na.rm = TRUE),
+    latest_reported_date = max(reported_date, na.rm = TRUE)
+  )
+```
+## A Note on Measurement
+
+-   gold data quality
+-   updates on funding/measurement changes + discontinuation
+-   likely many unreported cases
 

### 2024-09-24 scripts/Clean_Covid_Data.R

@@ -16,7 +16,7 @@ raw_covid_data <- read_csv("data/raw_data/raw_covid_data.csv")
 cleaned_covid_data <- raw_covid_data %>%
   
   ## Selecting the Variables of Interest
-  select(`X_id`, `Age.Group`, `Client.Gender`, `Reported.Date`, `Outcome`) %>%
+  select(`X_id`, `Age.Group`, `Client.Gender`, `Reported.Date`, `Ever.in.ICU`, `Ever.Hospitalized`, `Ever.Intubated`, `Outcome`)  %>%
   
   ## Cleaning the Names of the Variables
   janitor::clean_names() %>%

### 2024-09-24 scripts/Tests_Covid_Data.R

@@ -41,3 +41,7 @@ unique(cleaned_covid_data$age_group)
 # 3c) Test for Unique Outcome Values to only return RESOLVED, ACTIVE, and FATAL
 
 unique(cleaned_covid_data$outcome)
+
+## Test 4: Test to see if Reported Date is of type 'Date'
+
+class(cleaned_covid_data$reported_date)

### 2024-09-24 scripts/Tests_Covid_Data.R
Update Tests_Covid_Data.R

@@ -20,6 +20,7 @@ cleaned_covid_data <- read_csv("data/analysis_data/cleaned_covid_data.csv")
 sum(is.na(cleaned_covid_data$age_group))
 sum(is.na(cleaned_covid_data$client_gender))
 sum(is.na(cleaned_covid_data$outcome))
+sum(is.na(cleaned_covid_data$reported_date))
 
 
 ## Test 2: Empty Strings Test in Variables of Interest

### 2024-09-24 paper/paper.qmd
updates to data, paper,

@@
 
-### Table 1: Sample of Raw Covid-19 Data
+```{r}
+#| label: tbl-raw1
+#| tbl-cap: "COVID-19 Case Raw Data"
+#| echo: false
+#| warning: false
+#| message: false
 
-```{r, echo=FALSE, warning=FALSE, message=FALSE}
 library(kableExtra)
@@
   kable_styling(font_size = 8)
+
+```
+
+```{r}
+#| label: tbl-raw2
+#| tbl-cap: "COVID-19 Case Raw Data Part 2"
+#| echo: false
+#| warning: false
+#| message: false
 
@@
 
-### Table 2: Sample of Cleaned Covid-19 Data
-```{r, echo=FALSE, warning=FALSE, message=FALSE}
+```{r}
+#| label: tbl-clean
+#| tbl-cap: "COVID-19 Case Cleaned Data Part 2"
+#| echo: false
+#| warning: false
+#| message: false
+#| 
 # Cleaned Data Table (Table 2)
@@
 
-## Summary Statistics
-
-```{r, echo=FALSE, warning=FALSE, message=FALSE}
-cleaned_covid_data %>%
-  summarise(
-    total_cases = n(),
-    age_group_distribution = table(age_group),
-    gender_distribution = table(client_gender),
-    outcome_distribution = table(outcome),
-    earliest_reported_date = min(reported_date, na.rm = TRUE),
-    latest_reported_date = max(reported_date, na.rm = TRUE)
-  )
-```
 ## A Note on Measurement
@@
 
+@fig-cumulativecases displays the progression of COVID-19 cases in Toronto from January 2020 to March 2024.
+
+```{r}
+#| label: fig-cumulativecases
+#| fig-cap: "COVID-19 Case Progression Over Time, Toronto"
+#| fig.width: 5
+#| fig.height: 3
+#| fig.align: center
+#| echo: false
+#| warning: false
+#| message: false
+#| fig-pos: "H"
+
+# Load necessary libraries
+library(ggplot2)
+library(dplyr)
+
+# Step 1: Ensure reported_date is in Date format
+cleaned_covid_data$reported_date <- as.Date(cleaned_covid_data$reported_date)
+
+# Step 3: Plot cumulative cases over time
+ggplot(cleaned_covid_data, aes(x = reported_date, y = x_id)) +
+  geom_line(color = "blue", size = 1) +
+  labs(title = "Cumulative COVID-19 Cases Over Time in Toronto",
+       x = "Date",
+       y = "COVID-19 Cases") +
+  scale_x_date(limits = as.Date(c("2020-01-01", "2024-03-31")))
+
+```
+
+```{r}
+#| label: fig-treatmentoutcomes
+#| fig-cap: "Level of Hospitalisation vs Mortality Rates"
+#| fig.width: 5
+#| fig.height: 3
+#| fig.align: center
+#| echo: false
+#| warning: false
+#| message: false
+#| fig-pos: "H"
+
+# Step 1: Calculate mortality rate for intubated, hospitalized, and ICU patients
+# Create a summary table with mortality rates
+
+calc_mortality_rate <- function(df, condition_col) {
+  total_patients <- nrow(df %>% filter(!!sym(condition_col) == "Yes"))
+  fatal_patients <- nrow(df %>% filter(!!sym(condition_col) == "Yes" & outcome == "FATAL"))
+  if (total_patients > 0) {
+    return(fatal_patients / total_patients * 100)  # Return mortality rate as percentage
+  } else {
+    return(NA)  # Return NA if no patients in the group
+  }
+}
+
+# Apply the function for each group
+hospitalised_mortality_data <- data.frame(
+  Condition = c("Intubated", "Hospitalized", "ICU"),
+  Mortality_Rate = c(
+    calc_mortality_rate(cleaned_covid_data, "ever_intubated"),
+    calc_mortality_rate(cleaned_covid_data, "ever_hospitalized"),
+    calc_mortality_rate(cleaned_covid_data, "ever_in_icu")
+  )
+)
+
+# Step 2: Plot the mortality rate as a bar graph
+ggplot(hospitalised_mortality_data, aes(x = Condition, y = Mortality_Rate)) +
+  geom_bar(stat = "identity", fill = "steelblue") +
+  labs(title = "Mortality Rate by Condition (Intubated, Hospitalized, ICU)",
+       x = "Condition",
+       y = "Mortality Rate (%)") +
+  theme_minimal() +
+  ylim(0, 100)  # Set y-axis limits to 0-100 for percentage
+
+```
+
 ## Age vs Outcomes
 
+```{r}
+#| label: fig-agegroupmortality
+#| fig-cap: "Age Group vs Mortality Rates"
+#| fig.width: 5
+#| fig.height: 3
+#| fig.align: center
+#| echo: false
+#| warning: false
+#| message: false
+#| fig-pos: "H"
+
+# Step 1: Define the calc_mortality_rate function for age group
+calc_mortality_rate_age <- function(df, age_group_col) {
+  total_patients <- nrow(df %>% filter(!!sym(age_group_col) != ""))
+  fatal_patients <- nrow(df %>% filter(!!sym(age_group_col) != "" & outcome == "FATAL"))
+  if (total_patients > 0) {
+    return(fatal_patients / total_patients * 100)  # Return mortality rate as percentage
+  } else {
+    return(NA)  # Return NA if no patients in the group
+  }
+}
+
+# Step 2: Calculate mortality rate for each age group
+age_groups <- c("19 and younger", "20 to 29 Years", "30 to 39 Years", 
+                "40 to 49 Years", "50 to 59 Years", "60 to 69 Years", 
+                "70 to 79 Years", "80 to 89 Years", "90 and older")
+
+mortality_by_age_group <- data.frame(
+  Age_Group = age_groups,
+  Mortality_Rate = sapply(age_groups, function(group) {
+    df_age_group <- cleaned_covid_data %>% filter(age_group == group)
+    calc_mortality_rate_age(df_age_group, "age_group")
+  })
+)
+
+# Step 3: Plot Age Group vs Mortality Rate as a bar chart
+ggplot(mortality_by_age_group, aes(x = Age_Group, y = Mortality_Rate)) +
+  geom_bar(stat = "identity", fill = "steelblue") +
+  labs(title = "Mortality Rate by Age Group",
+       x = "Age Group",
+       y = "Mortality Rate (%)") +
+  theme_minimal() +
+  ylim(0, 15) +  # Set y-axis limits to 0-15 for percentage
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability
+
+```
+
+```{r}
+#| label: fig-agegrouphospitalisation
+#| fig-cap: "Chance of Hospitalisation by Age Group"
+#| fig.width: 5
+#| fig.height: 3
+#| fig.align: center
+#| echo: false
+#| warning: false
+#| message: false
+#| fig-pos: "H"
+
+# Step 1: Define a function to calculate rates for each condition (hospitalization, intubation, ICU)
+calc_condition_rate <- function(df, condition_col) {
+  total_patients <- nrow(df %>% filter(!is.na(!!sym(condition_col))))  # Non-missing values for the condition
+  condition_patients <- nrow(df %>% filter(!!sym(condition_col) == "Yes"))
+  if (total_patients > 0) {
+    return(condition_patients / total_patients * 100)  # Return rate as a percentage
+  } else {
+    return(NA)  # Return NA if no patients in the group
+  }
+}
+
+# Step 2: Calculate the rates for hospitalization, intubation, and ICU admission by age group
+age_groups <- c("19 and younger", "20 to 29 Years", "30 to 39 Years", 
+                "40 to 49 Years", "50 to 59 Years", "60 to 69 Years", 
+                "70 to 79 Years", "80 to 89 Years", "90 and older")
+
+hospitalisation_data <- data.frame(
+  Age_Group = age_groups,
+  Hospitalization_Rate = sapply(age_groups, function(group) {
+    df_age_group <- cleaned_covid_data %>% filter(age_group == group)
+    calc_condition_rate(df_age_group, "ever_hospitalized")
+  }),
+  Intubation_Rate = sapply(age_groups, function(group) {
+    df_age_group <- cleaned_covid_data %>% filter(age_group == group)
+    calc_condition_rate(df_age_group, "ever_intubated")
+  }),
+  ICU_Rate = sapply(age_groups, function(group) {
+    df_age_group <- cleaned_covid_data %>% filter(age_group == group)
+    calc_condition_rate(df_age_group, "ever_in_icu")
+  })
+)
+
+# Step 3: Reshape the data to long format for grouped bar plot
+long_hosp_data <- hospitalisation_data %>%
+  pivot_longer(cols = c(Hospitalization_Rate, Intubation_Rate, ICU_Rate),
+               names_to = "Condition",
+               values_to = "Rate")
+
+# Step 4: Plot Age Group vs Chance of Hospitaliation, Intubation, or ICU as a grouped bar chart
+ggplot(long_hosp_data, aes(x = Age_Group, y = Rate, fill = Condition)) +
+  geom_bar(stat = "identity", position = "dodge") +
+  labs(title = "Hospitalisation, Intubation, ICU Admission by Age Group",
+       x = "Age Group",
+       y = "Percentage (%)") +
+  theme_minimal() +
+  ylim(0, 25) +  # Set y-axis limits to 0-100%
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability
+```
 ## Gender vs Outcomes
 
+```{r}
+#| label: fig-gendermortality
+#| fig-cap: "COVID-19 Mortality by Gender"
+#| fig.width: 5
+#| fig.height: 3
+#| fig.align: center
+#| echo: false
+#| warning: false
+#| message: false
+#| fig-pos: "H"
+
+# Step 1: Define the calc_mortality_rate function (as used previously)
+calc_mortality_rate_gender <- function(df, gender_col) {
+  total_patients <- nrow(df %>% filter(!!sym(gender_col) != ""))
+  fatal_patients <- nrow(df %>% filter(!!sym(gender_col) != "" & outcome == "FATAL"))
+  if (total_patients > 0) {
+    return(fatal_patients / total_patients * 100)  # Return mortality rate as percentage
+  } else {
+    return(NA)  # Return NA if no patients in the group
+  }
+}
+
+# Step 2: Calculate mortality rate for each gender (MALE, FEMALE, OTHER)
+genders <- c("MALE", "FEMALE", "OTHER")
+
+mortality_by_gender <- data.frame(
+  Client_Gender = genders,
+  Mortality_Rate = sapply(genders, function(gender) {
+    df_gender <- cleaned_covid_data %>% filter(client_gender == gender)
+    calc_mortality_rate_gender(df_gender, "client_gender")
+  })
+)
+
+# Step 3: Plot Gender vs Mortality Rate as a bar chart
+ggplot(mortality_by_gender, aes(x = Client_Gender, y = Mortality_Rate)) +
+  geom_bar(stat = "identity", fill = "steelblue") +
+  labs(title = "Mortality Rate by Gender",
+       x = "Client Gender",
+       y = "Mortality Rate (%)") +
+  theme_minimal() +
+  ylim(0, 2)  # Set y-axis limits to 0-100 for percentage
+
+```
 # Discussion {#sec-discussion}
@@
 
+# LLM Disclosure
+
 # References
-
-# LLM Disclosure

### 2024-09-24 scripts/Clean_Covid_Data.R

@@ -41,3 +41,4 @@ cleaned_covid_data <- raw_covid_data %>%
 
 #### Save data ####
 write_csv(cleaned_covid_data, "data/analysis_data/cleaned_covid_data.csv")
+

### 2024-09-24 paper/paper.qmd
updating csv files and writing the paper

@@
 
-(\scriptsize Table Separated using Tips from [@stack_tip] and LLMs)
+(Table Separated using Tips from [@stack_tip] and LLMs)
 
@@
 
+\scriptsize
 ```{r}
@@
 ```
+\normalsize
 
@@
 
+\newpage
+
 # Results {#sec-results}
 
-## Cases over Time
+## General Results
 
 @fig-cumulativecases displays the progression of COVID-19 cases in Toronto from January 2020 to March 2024.
+@fig-treatmentoutcomes importantly highlights mortality rates by level of hospitalisation.
+That is, the mortality rates for those hospitalised, put in the ICU, and/or intubated.
 
@@
 #| fig.width: 5
-#| fig.height: 3
+#| fig.height: 2.75
 #| fig.align: center
@@
   geom_line(color = "blue", size = 1) +
-  labs(title = "Cumulative COVID-19 Cases Over Time in Toronto",
+  labs(title = "COVID-19 Cases Over Time, Toronto",
        x = "Date",
        y = "COVID-19 Cases") +
+  theme(
+    plot.title = element_text(size = 12),        # Reduce the title font size
+    axis.title.x = element_text(size = 10),      # Reduce the x-axis title font size
+    axis.title.y = element_text(size = 10),      # Reduce the y-axis title font size
+    axis.text.x = element_text(size = 8),        # Reduce the x-axis text (labels) font size
+    axis.text.y = element_text(size = 8),        # Reduce the y-axis text (labels) font size
+  ) +
   scale_x_date(limits = as.Date(c("2020-01-01", "2024-03-31")))
@@
   geom_bar(stat = "identity", fill = "steelblue") +
-  labs(title = "Mortality Rate by Condition (Intubated, Hospitalized, ICU)",
+  labs(title = "Mortality by Condition (Intubated, Hospitalized, ICU)",
        x = "Condition",
@@
   theme_minimal() +
+  theme(
+    plot.title = element_text(size = 12),        # Reduce the title font size
+    axis.title.x = element_text(size = 10),      # Reduce the x-axis title font size
+    axis.title.y = element_text(size = 10),      # Reduce the y-axis title font size
+    axis.text.x = element_text(size = 8),        # Reduce the x-axis text (labels) font size
+    axis.text.y = element_text(size = 8),        # Reduce the y-axis text (labels) font size
+  ) +
   ylim(0, 100)  # Set y-axis limits to 0-100 for percentage
@@
 
+@fig-agegroupmortality shows the differences in mortality rates for different age groups at the time of contracting COVID-19.
+@fig-agegrouphospitalisation shows the hospitalisation, ICU, or intubation rates for different age groups.
 ```{r}
@@
 #| label: fig-agegrouphospitalisation
-#| fig-cap: "Chance of Hospitalisation by Age Group"
+#| fig-cap: "Hospitalisation, Intubation, ICU Admission Rate by Age"
 #| fig.width: 5
@@
   geom_bar(stat = "identity", position = "dodge") +
-  labs(title = "Hospitalisation, Intubation, ICU Admission by Age Group",
+  labs(title = "Hospitalisation, Intubation, ICU Admission Rate by Age",
        x = "Age Group",
@@
   theme_minimal() +
+  theme(
+    plot.title = element_text(size = 12),        # Reduce the title font size
+    axis.title.x = element_text(size = 10),      # Reduce the x-axis title font size
+    axis.title.y = element_text(size = 10),      # Reduce the y-axis title font size
+    axis.text.x = element_text(size = 8),        # Reduce the x-axis text (labels) font size
+    axis.text.y = element_text(size = 8),        # Reduce the y-axis text (labels) font size
+    legend.title = element_text(size = 8),       # Reduce the legend title font size
+    legend.text = element_text(size = 6)         # Reduce the legend text font size
+  ) +
   ylim(0, 25) +  # Set y-axis limits to 0-100%
@@
 ```
+\newpage
+
 ## Gender vs Outcomes
 
+@fig-gendermortality graphs COVID-19 mortality rates by gender.
+@fig-genderhospitalisation graphs hospitalisation, ICU and/or incubation rates by gender.
+That is, how likely a person who has contracted COVID-19 is likely to be in any of these conditions, by gender.
 ```{r}
@@
 ```
+
+```{r}
+#| label: fig-genderhospitalisation
+#| fig-cap: "Hospitalisation, ICU, Intubation Rate by Gender"
+#| fig.width: 5
+#| fig.height: 3
+#| fig.align: center
+#| echo: false
+#| warning: false
+#| message: false
+#| fig-pos: "H"
+# Step 1: Calculate the rates for hospitalization, intubation, and ICU admission by gender
+genders <- c("MALE", "FEMALE", "OTHER")
+
+hospitalization_data_gender <- data.frame(
+  Client_Gender = genders,
+  Hospitalization_Rate = sapply(genders, function(gender) {
+    df_gender <- cleaned_covid_data %>% filter(client_gender == gender)
+    calc_condition_rate(df_gender, "ever_hospitalized")
+  }),
+  Intubation_Rate = sapply(genders, function(gender) {
+    df_gender <- cleaned_covid_data %>% filter(client_gender == gender)
+    calc_condition_rate(df_gender, "ever_intubated")
+  }),
+  ICU_Rate = sapply(genders, function(gender) {
+    df_gender <- cleaned_covid_data %>% filter(client_gender == gender)
+    calc_condition_rate(df_gender, "ever_in_icu")
+  })
+)
+
+# Step 2: Reshape the data to long format for grouped bar plot
+long_data_gender <- hospitalization_data_gender %>%
+  pivot_longer(cols = c(Hospitalization_Rate, Intubation_Rate, ICU_Rate),
+               names_to = "Condition",
+               values_to = "Rate")
+
+# Step 3: Plot Gender vs Chance of Hospitalization, Intubation, or ICU as a grouped bar chart
+ggplot(long_data_gender, aes(x = Client_Gender, y = Rate, fill = Condition)) +
+  geom_bar(stat = "identity", position = "dodge") +
+  labs(title = "Hospitalization, Intubation, or ICU by Gender",
+       x = "Gender",
+       y = "Percentage (%)") +
+  theme_minimal() +
+  theme(
+    plot.title = element_text(size = 12),        # Reduce the title font size
+    axis.title.x = element_text(size = 10),      # Reduce the x-axis title font size
+    axis.title.y = element_text(size = 10),      # Reduce the y-axis title font size
+    axis.text.x = element_text(size = 8),        # Reduce the x-axis text (labels) font size
+    axis.text.y = element_text(size = 8),        # Reduce the y-axis text (labels) font size
+    legend.title = element_text(size = 8),       # Reduce the legend title font size
+    legend.text = element_text(size = 6)         # Reduce the legend text font size
+  ) +
+  ylim(0, 10) +  # Set y-axis limits to 0-100%
+  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))  # Keep x-axis labels horizontal
+```
 # Discussion {#sec-discussion}
@@
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+Understanding the effect age has on COVID-19 is extremely important both for potential victims and those in charge of policy to put in place correct and appropriate health measures for different ages.
+
+Looking at @fig-agegroupmortality, we see a clear message - age group and mortality are positively correlated.
+More specifically, mortality past the ages of 50-59 shoots from under half a percent to 1.72% for 60-69 years of age.
+This only worsens as individuals get older.
+For the age groups 70-79 and 80-89, the mortality rate if one contracts COVID-19 shoots to 4.14% and 7.50% respectively.
+Moreover, those who contract COVID-19 at ages of 90 and above have a mortality rate of almost 10% (9.62%).
+
+In contrast, those younger than 50 at the time of contraction have mortality rates under 0.5%.
+In fact, those 19 and younger have a mortality rate of only 0.007%.
+Clearly, age and mortalty are very strongly correlated.
+
+Perhaps of more relevance are chances of hospitalisation, admission to the ICU, and intubation.
+In reference to @fig-agegrouphospitalisation, we see that the older population is much more susceptible to having more severe cases of COVID-19.
+For all ages above 70, the chance of hospitalisation is above 15%, peaking at 19.46% for 80-89 year olds.
+Compared to the younger population, this is a stark difference.
+Those under 30 have hospitalisation rates of under 30%.
+This increases with age: hospital admission rates for those with COVID-19 is 1.15% for 30-39 year olds, 2.13% for 40-49 year olds, 4.15% for 50-59 year olds, and 8.91% for 60-69 year olds.
+
+discussion about weaker immune systems with age (bring in some literature review)
 
@@
 
+Similar to age, understanding differences in outcomes for genders matter greatly to virologists and as a matter of public safety.
+Looking at @fig-gendermortality, we find that males that contract COVID-19 have a much higher mortality rate when compared to females (1.47% vs 1.06%).
+According to this data, males who contract COVID-19 are almost 1.4 times more likely to die than females.
+
+Similarly, looking at @fig-genderhospitalisation, we see that males are more likely to be hospitalised, put in the ICU, and intubated.
+
+discussion about how Covid-19 affects the male body vs female; lit and scientific review; is this a worldwide trend?
+
 ## Broader Discussion and Takeaways
+
+Reference the importance of @fig-treatmentoutcomes - those intubated have almost a 50% chance of death. (what?!)
 
@@
 
-Weaknesses and next steps should also be included.
+Measurement weaknesses (stopped collecting data), might be important to look at selection biases (are there more male cases than females? what about Other genders? are they affect results?
 
@@
 
+ChatGPT Data Analyst was used to generate code and help fix bugs for this assignment. A full LLM Disclosure can be found on the GitHub Repository under "Other - LLM - usage.txt"
+
 # References

### 2024-09-24 paper/paper.qmd
updates to abstract

@@
 ---
-title: "Toronto's Covid-19 Impact: Effect of Age and Gender on Outcomes"
+title: "Toronto's COVID-19 Impact: Effect of Age and Gender on Outcomes"
 author: 
@@
 date-format: long
-abstract: "The Covid-19 Pandemic caused irreparable damage to the infrastructure of our global systems, highlighting the neccessity to better understand factors that influence health outcomes.
+abstract: "The COVID-19 Pandemic caused irreparable damage to the infrastructure of our global systems, highlighting the neccessity to better understand factors that influence health outcomes.
 This paper uses data from OpenDataToronto to investigate how age and gender affects outcomes of those infected with the virus.
-xyz about results"
+We find that in Toronto, age is strongly correlated with mortality and poorer hospitalisation outcomes.
+Further, we find that males in Toronto are almost 1.4 times more likely to die from COVID-19 than females."
 format: pdf

### 2024-09-24 paper/paper.qmd
changes to paper and readme

@@
 We find that in Toronto, age is strongly correlated with mortality and poorer hospitalisation outcomes.
-Further, we find that males in Toronto are almost 1.4 times more likely to die from COVID-19 than females."
+Further, we find that males have significantly worse outcomes than females in both mortality rates and hopsitalisation outcomes"
 format: pdf

### 2024-09-27 paper/paper.qmd
updates to paper and references file

@@
 #Loading Packages
-library(pander)
 library(tidyverse)
+library(lintr)
 library(here)
@@
 
-\newpage
-
 # Introduction
@@
 As of April 13, 2024, Worldometer Info estimates that over 704 million people worldwide have contracted the virus, resulting in approximately 7 million deaths [@worldometer].
-In Canada alone, there have been almost 4.6 million reported cases and more than 38,000 deaths as of July 20, 2024 [@infobase_ca].
-
-While these statistics paint a harrowing picture of the human toll inflicted, the pandemic has exacerbated existing economic inequalities, destabilized political systems, and put immense pressure on societal infrastructure globally.
+In Canada alone, there have been almost 4.6 million reported cases and more than 38,000 deaths as of July 20, 2024 [@infobase_ca].While these statistics paint a harrowing picture of the human toll inflicted, the pandemic has exacerbated existing economic inequalities, destabilized political systems, and put immense pressure on societal infrastructure globally.
 Looking past the immediate health crisis, lockdowns and restrictions have had immeasurable impacts not only on global supply chains, but also on the mental heatlh of many forced to quarantine or self isolate.
 
-An article by Fortune valued the economic burden on the US Economy to be upwards of \$14 Trillion USD at the end of 2023 [@fortune].
-While this figure mainly took into account the 'standard economic effects' of the pandemic - revenue lost due to mandatory business closures, decreases in air travel, and workplace absences - the article noted that there were many unobservable factors that were incredibly burdernsome to the economy, such as long term physical and mental health effects of the pandemic on the population.
-
-Covid-19 has been found to have varying effects across demographics.
+Importantly, Covid-19 has been found to have varying effects across demographics.
 A 2020 article published in the PLOS Journal found that 'Covid-19 may be associated with worse outcomes in males than in females'.
 The article found that men are up to 22% more likely to require ICU admission.
-
-Moreover, an article published in the Springer Link Journal in 2021 found that older adults, in particular those above the age of 65, face higher mortality rates than their younger counterparts.
+@plos Moreover, an article published in the Springer Link Journal in 2021 found that older adults, in particular those above the age of 65, face higher mortality rates than their younger counterparts.
 Weaker immune systems, and the higher likely presence of other conditions can exacerbate the effects of the virus.
+@springer
 
@@
 
-The remainder of this paper is structured as follows. @sec-data discusses @sec-results xyz @sec-discussion xyz.
+The remainder of this paper is structured as follows.
+@sec-data discusses data selecting and cleaning techniques, and provides a sample of the raw and cleaned data.
+@sec-results displays relevant figures found through the analysis - general results, age specific results, and gender specific results.
+@sec-discussion discusses these results in detail, the broader significance of the findings, and highlights some weaknesses and next steps that could be taken to improve the report.
 
@@
 
-We use @citeR and @rohan.
+## A Note on Data Selection and Measurement
 
-## Data Selection
+Data used in this report was sourced from OpenDataToronto's portal. @opendatatoronto. More specifically, the data set "Covid-19 Cases in Toronto" was used and cleaned for the purposes of this report.
 
-Data used in this report was sourced from OpenDataToronto's portal.
-@opendatatoronto.
-More specifically, the dataset "Covid-19 Cases in Toronto" was used and cleaned for the purposes of this report.
-Toronto Public Health (TPH) released anonymized, person-level data from the start of the pandemic in January 2020.
-The data spans from the first reported case on in January of 2020 to February 14th, 2024.
+Toronto Public Health (TPH) collects COVID-19 case data in different ways.
+If an individual tests positive at a healthcare provider or laboratory, it is reported to TPH.
+Furthermore, in 'high-risk settings', such as hospitals and nursing homes, management is required to report positive cases to TPH.
+TPH also uses extensive contact tracing methods to monitor potential spreads of COVID-19.
+@covid_reporting.
+They released anonymized, person-level data from the start of the pandemic in January 2020 to  OpenDataToronto.The data spans from the first reported case on in January of 2020 to February 14th, 2024.
 In a statement on the website for this dataset, OpenDataToronto states that "As case and outbreak management guidelines changed and COVID-19 specific resources were no longer funded, the level of detail available for cases decreased, and more recent data are less complete and not comparable to previous years.
-TPH discontinued the production of this report with the final refresh as of February 14, 2024" @opendatatoronto_set
+TPH discontinued the production of this report with the final refresh as of February 14, 2024" @opendatatoronto_set.
+
+The data set received a 'Gold' quality score, which takes into account the freshness, usability, metadata, accessibility and completeness of the data.
+This adds to the credibility of the data.
+However, there are likely many unreported cases due to the fact that it is not mandatory to report COVID-19 if you test positive using an over-the-counter antigen test.
+This likely leads to some under-reporting, and hence the data might not be fully representative of the overall effect COVID-19 had in Toronto.
+
+This project uses @citeR to simulate, download, clean, plot, and analyse the data provided by @opendatatoronto-set.
 
@@
 In it's original form, the dataset contains more than 414,000 entries regarding information on cases of Covid-19 in Toronto.
-Below are two tables with the first three rwos of the raw data, separated into two tables for readability.
+@tbl-raw1 and @tbl-raw2 below are samples of the first three rows of the raw data, separated into two tables for readability.
 
@@
 #| label: tbl-raw1
-#| tbl-cap: "COVID-19 Case Raw Data"
+#| tbl-cap: "COVID-19 Case Raw Data (Part 1)"
 #| echo: false
@@
 raw_covid_data %>%
-  select(X_id, Outbreak.Associated, Assigned_ID, Age.Group, Neighbourhood.Name, FSA, Source.of.Infection, Classification) %>%
+  select(X_id, Outbreak.Associated, Age.Group, Neighbourhood.Name, FSA, Source.of.Infection, Classification) %>%
   head(3) %>%
@@
 #| label: tbl-raw2
-#| tbl-cap: "COVID-19 Case Raw Data Part 2"
+#| tbl-cap: "COVID-19 Case Raw Data (Part 2)"
 #| echo: false
@@
 
+There were a variety of issues and fixes needed to be done to the raw data in order to get it ready for analysis.
+In order to download, clean and run tests on the data, the package Tidyverse, @rohan was used.
+
+When cleaning the data, the first step was to select appropriate variables of interest to analyse.
+The raw data had many different variables to pick from, but this report importantly selected the variables 'Age Group', 'Client Gender', 'Reported Date', and 'Outcome'.
+This was done to run some basic analysis for the differences in outcomes for age groups and genders.
+Additionally, the variables 'Ever Hospitalized', 'Ever In Icu', and 'Ever Intubated' were selected to provide further depth to analysis regarding the severity of cases across age and gender.
+
+Next, the Janitor package as part of Tidyverse was used to clean up the names of the variables, converting them to snake lowercase format; for example, 'Age.Group' was cleaned to 'age_group'.
+Another key step in the cleaning process was to arrange the data in chronological order, and then match Case ID's to the chronologically adjusted dates.
+This was done in order to have a clearer picture on the time trends of mortality and case progression of COVID-19 in Toronto.
+Further, the variable 'Client Gender' took on many different values in the raw data, so for simplicity, genders other than Male and Female were combined to take the value 'Other'.
+Finally, NA values were removed.
+After running tests, there were 422 NA values associated with the raw data, which were dropped.
+
+@tbl-clean is a sample of the first 3 rows of the new cleaned data table.
 \scriptsize
@@
 #| label: tbl-clean
-#| tbl-cap: "COVID-19 Case Cleaned Data Part 2"
+#| tbl-cap: "Sample of COVID-19 Case Cleaned Data"
 #| echo: false
@@
 #| 
-# Cleaned Data Table (Table 2)
+# Cleaned Data Table
 cleaned_covid_data %>%
-  head(5) %>%
+  head(3) %>%
   kable(col.names = c("x_id", "age_group", "client_gender", "reported_date", "ever_in_icu", "ever_hospitalized", "ever_intubated", "outcome"),
@@
 \normalsize
-
-## A Note on Measurement
-
--   gold data quality
--   updates on funding/measurement changes + discontinuation
--   likely many unreported cases
 
@@
 ```
-\newpage
+
+```{r}
+#| label: fig-agegroup-mortality-time-line
+#| fig-cap: "Mortality Rate Over Time by Age Group"
+#| fig.width: 7
+#| fig.height: 4
+#| fig.align: center
+#| echo: false
+#| warning: false
+#| message: false
+
+# Step 1: Convert reported_date to Date format if not already done
+cleaned_covid_data$reported_date <- as.Date(cleaned_covid_data$reported_date)
+
+# Step 2: Aggregate data by age group and month, calculate mortality rate for each month
+mortality_by_age_group_month <- cleaned_covid_data %>%
+  mutate(month = floor_date(reported_date, "month")) %>%  # Create a new column for month
+  group_by(age_group, month) %>%
+  summarize(
+    total_cases = n(),
+    fatal_cases = sum(outcome == "FATAL"),
+    mortality_rate = ifelse(total_cases > 0, fatal_cases / total_cases * 100, NA)
+  ) %>%
+  filter(!is.na(mortality_rate))  # Remove rows where mortality_rate is NA
+
+# Custom color palette for age groups
+age_group_colors <- c(
+  "19 and younger" = "#1f77b4",  # Blue
+  "20 to 29 Years" = "#ff7f0e",  # Orange
+  "30 to 39 Years" = "#2ca02c",  # Green
+  "40 to 49 Years" = "#d62728",  # Red
+  "50 to 59 Years" = "#9467bd",  # Purple
+  "60 to 69 Years" = "#8c564b",  # Brown
+  "70 to 79 Years" = "#e377c2",  # Pink
+  "80 to 89 Years" = "#17becf",  # Cyan
+  "90 and older" = "#bcbd22"     # Yellow-green
+)
+
+# Step 3: Plot the mortality rate over time (monthly aggregation) with custom colors
+ggplot(mortality_by_age_group_month, aes(x = month, y = mortality_rate, color = age_group)) +
+  geom_line(size = 1) +  # Create a line plot for each age group
+  scale_color_manual(values = age_group_colors) +  # Apply custom colors
+  labs(
+    title = "Mortality Rate Over Time by Age Group (Monthly Aggregated Data)",
+    x = "Date",
+    y = "Mortality Rate (%)"
+  ) +
+  theme_minimal() +
+  theme(
+    plot.title = element_text(size = 14),
+    axis.title.x = element_text(size = 12),
+    axis.title.y = element_text(size = 12),
+    legend.title = element_text(size = 10),
+    legend.text = element_text(size = 8)
+  )
+```
 
@@
 ```
+
+```{r}
+#| label: fig-gender-mortality-time-line
+#| fig-cap: "Mortality Rate Over Time by Gender"
+#| fig.width: 7
+#| fig.height: 4
+#| fig.align: center
+#| echo: false
+#| warning: false
+#| message: false
+
+# Step 1: Ensure reported_date is in Date format if not already done
+cleaned_covid_data$reported_date <- as.Date(cleaned_covid_data$reported_date)
+
+# Step 2: Aggregate data by gender and month, calculate mortality rate for each month
+mortality_by_gender_month <- cleaned_covid_data %>%
+  mutate(month = floor_date(reported_date, "month")) %>%  # Create a new column for month
+  group_by(client_gender, month) %>%
+  summarize(
+    total_cases = n(),
+    fatal_cases = sum(outcome == "FATAL"),
+    mortality_rate = ifelse(total_cases > 0, fatal_cases / total_cases * 100, NA)
+  ) %>%
+  filter(!is.na(mortality_rate))  # Remove rows where mortality_rate is NA
+
+# Custom color palette for genders
+gender_colors <- c(
+  "MALE" = "#1f77b4",   # Blue
+  "FEMALE" = "#ff7f0e", # Orange
+  "OTHER" = "#2ca02c"   # Green
+)
+
+# Step 3: Plot the mortality rate over time (monthly aggregation) by gender with custom colors
+ggplot(mortality_by_gender_month, aes(x = month, y = mortality_rate, color = client_gender)) +
+  geom_line(size = 1) +  # Create a line plot for each gender
+  scale_color_manual(values = gender_colors) +  # Apply custom colors
+  labs(
+    title = "Mortality Rate Over Time by Gender (Monthly Aggregated Data)",
+    x = "Date",
+    y = "Mortality Rate (%)",
+    color = "Gender"
+  ) +
+  theme_minimal() +
+  theme(
+    plot.title = element_text(size = 14),
+    axis.title.x = element_text(size = 12),
+    axis.title.y = element_text(size = 12),
+    legend.title = element_text(size = 10),
+    legend.text = element_text(size = 8)
+  )
+```
 # Discussion {#sec-discussion}
@@
 
-\newpage
-
 # LLM Disclosure

### 2024-09-27 scripts/Simulate_Covid_Data.R
updates to simulate script

@@ -10,7 +10,6 @@
 
 #### Workspace setup ####
 library(tidyverse)
-library(dplyr)
 
 #### Simulate data ####
 

### 2024-09-27 paper/paper.qmd
Final changes to paper and references file

@@
 We find that in Toronto, age is strongly correlated with mortality and poorer hospitalisation outcomes.
-Further, we find that males have significantly worse outcomes than females in both mortality rates and hopsitalisation outcomes"
+Further, we find that males have significantly worse outcomes than females in both mortality rates and hopsitalisation outcomes."
 format: pdf
@@
 A 2020 article published in the PLOS Journal found that 'Covid-19 may be associated with worse outcomes in males than in females'.
-The article found that men are up to 22% more likely to require ICU admission.
-@plos Moreover, an article published in the Springer Link Journal in 2021 found that older adults, in particular those above the age of 65, face higher mortality rates than their younger counterparts.
-Weaker immune systems, and the higher likely presence of other conditions can exacerbate the effects of the virus.
-@springer
+The article found that men are up to 22% more likely to require ICU admission [@plos].
+Moreover, an article published in the Springer Link Journal in 2021 found that older adults, in particular those above the age of 65, face higher mortality rates than their younger counterparts.
+Weaker immune systems, and the higher likely presence of other conditions can exacerbate the effects of the virus [@springer].
 
@@
 
-Data used in this report was sourced from OpenDataToronto's portal. @opendatatoronto. More specifically, the data set "Covid-19 Cases in Toronto" was used and cleaned for the purposes of this report.
+Data used in this report was sourced from OpenDataToronto's portal [@opendatatoronto]. More specifically, the data set "COVID-19 Cases in Toronto" was used and cleaned for the purposes of this report.
 
@@
 Furthermore, in 'high-risk settings', such as hospitals and nursing homes, management is required to report positive cases to TPH.
-TPH also uses extensive contact tracing methods to monitor potential spreads of COVID-19.
-@covid_reporting.
-They released anonymized, person-level data from the start of the pandemic in January 2020 to  OpenDataToronto.The data spans from the first reported case on in January of 2020 to February 14th, 2024.
+TPH also uses extensive contact tracing methods to monitor potential spreads of COVID-19 [@covid_reporting].
+They released anonymizsd, person-level data from the start of the pandemic in January 2020 to  OpenDataToronto.The data spans from the first reported case on in January of 2020 to February 14th, 2024.
 In a statement on the website for this dataset, OpenDataToronto states that "As case and outbreak management guidelines changed and COVID-19 specific resources were no longer funded, the level of detail available for cases decreased, and more recent data are less complete and not comparable to previous years.
-TPH discontinued the production of this report with the final refresh as of February 14, 2024" @opendatatoronto_set.
+TPH discontinued the production of this report with the final refresh as of February 14, 2024" [@opendatatoronto_set].
 
@@
 However, there are likely many unreported cases due to the fact that it is not mandatory to report COVID-19 if you test positive using an over-the-counter antigen test.
-This likely leads to some under-reporting, and hence the data might not be fully representative of the overall effect COVID-19 had in Toronto.
+This likely leads to some under-reporting, and hence the data might not be fully representative of the overall impact COVID-19 had in Toronto.
 
-This project uses @citeR to simulate, download, clean, plot, and analyse the data provided by @opendatatoronto-set.
+This project uses @citeR to simulate, download, clean, plot, and analyse the data provided by @opendatatoronto_set.
 
@@
 There were a variety of issues and fixes needed to be done to the raw data in order to get it ready for analysis.
-In order to download, clean and run tests on the data, the package Tidyverse, @rohan was used.
+In order to download, clean and run tests on the data, the package Tidyverse, [@rohan] was used.
 
@@
 @fig-cumulativecases displays the progression of COVID-19 cases in Toronto from January 2020 to March 2024.
-@fig-treatmentoutcomes importantly highlights mortality rates by level of hospitalisation.
-That is, the mortality rates for those hospitalised, put in the ICU, and/or intubated.
+@fig-treatmentoutcomes importantly shows mortality rates by the severity of hospitalisation outcomes.
 
@@
 #| fig.width: 5
-#| fig.height: 2.75
+#| fig.height: 2.85
 #| fig.align: center
@@
        x = "Date",
-       y = "COVID-19 Cases") +
+       y = "COVID-19 Cases (Hundred Thousands)") +
   theme(
@@
   ) +
+  scale_y_continuous(labels = scales::label_number(scale = 1e-5)) +
   scale_x_date(limits = as.Date(c("2020-01-01", "2024-03-31")))
@@
 
-@fig-agegroupmortality shows the differences in mortality rates for different age groups at the time of contracting COVID-19.
-@fig-agegrouphospitalisation shows the hospitalisation, ICU, or intubation rates for different age groups.
+@fig-agegroupmortality shows differences in mortality rates for different age groups at the time of contracting COVID-19. 
 ```{r}
@@
 #| fig-cap: "Age Group vs Mortality Rates"
-#| fig.width: 5
-#| fig.height: 3
+#| fig.width: 4.75
+#| fig.height: 2.8
 #| fig.align: center
@@
   theme_minimal() +
-  ylim(0, 15) +  # Set y-axis limits to 0-15 for percentage
+  ylim(0, 10) +  # Set y-axis limits to 0-15 for percentage
   theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability
@@
 ```
+@fig-fatalities-pie shows the share of fatalities made up by larger binned age groups.
+```{r}
+#| label: fig-fatalities-pie
+#| fig-cap: "Fatalities by Age Group"
+#| fig.width: 5
+#| fig.height: 3
+#| fig.align: center
+#| echo: false
+#| warning: false
+#| message: false
+#| fig-pos: "H"
 
+# Step 1: Group age groups as required
+cleaned_covid_data <- cleaned_covid_data %>%
+  mutate(age_group_binned = case_when(
+    age_group %in% c("60 to 69 Years", "70 to 79 Years", "80 to 89 Years", "90 and older") ~ "60+",
+    age_group %in% c("30 to 39 Years", "40 to 49 Years", "50 to 59 Years") ~ "30-59",
+    age_group %in% c("19 and younger", "20 to 29 Years") ~ "Under 30",
+    TRUE ~ "Unknown"
+  ))
+
+# Step 2: Filter only fatal outcomes and calculate percentages
+fatalities_by_age_group <- cleaned_covid_data %>%
+  filter(outcome == "FATAL") %>%
+  group_by(age_group_binned) %>%
+  summarise(total_fatalities = n()) %>%
+  ungroup() %>%
+  mutate(percentage = total_fatalities / sum(total_fatalities) * 100)  # Calculate percentage
+
+# Step 3: Create a pie chart with annotations for "60+"
+ggplot(fatalities_by_age_group, aes(x = "", y = total_fatalities, fill = age_group_binned)) +
+  geom_bar(stat = "identity", width = 1) +
+  coord_polar("y", start = 0) +
+  labs(
+    title = "Fatalities by Age Group",
+    x = NULL,
+    y = NULL,
+    fill = "Age Group"  # Set legend title to "Age Group"
+  ) +
+  theme_void() +  # Remove background, gridlines, and axis labels
+  theme(
+    plot.title = element_text(size = 14, hjust = 0.5),  # Center the title
+    legend.title = element_text(size = 12),  # Adjust the legend title size
+    legend.text = element_text(size = 10),   # Adjust the legend text size
+    legend.position = "right"                # Position the legend on the right
+  ) +
+  geom_text(
+    data = fatalities_by_age_group %>% filter(age_group_binned %in% c("60+")),  # Annotate only 60+
+    aes(label = paste0(round(percentage, 1), "%")),  # Add percentage label
+    position = position_stack(vjust = 0.5),  # Place text inside the pie chart
+    size = 5,  # Font size of the annotation
+    color = "black"  # Text color
+  ) +
+  scale_fill_manual(values = c("60+" = "skyblue", "30-59" = "orange", "Under 30" = "green", "Unknown" = "gray"))
+```
+@fig-agegrouphospitalisation shows the hospitalisation, ICU, and intubation rates for different age groups.
 ```{r}
@@
 #| fig.width: 5
-#| fig.height: 3
+#| fig.height: 2.9
 #| fig.align: center
@@
 ```
-
+@fig-agetimemortality shows how mortality rates have progressed over time for different age groups.
 ```{r}
-#| label: fig-agegroup-mortality-time-line
+#| label: fig-agetimemortality
 #| fig-cap: "Mortality Rate Over Time by Age Group"
-#| fig.width: 7
-#| fig.height: 4
+#| fig.width: 6.8
+#| fig.height: 3.75
 #| fig.align: center
@@
 #| message: false
+#| fig-pos: "H"
 
@@
 
-@fig-gendermortality graphs COVID-19 mortality rates by gender.
-@fig-genderhospitalisation graphs hospitalisation, ICU and/or incubation rates by gender.
-That is, how likely a person who has contracted COVID-19 is likely to be in any of these conditions, by gender.
+@fig-gendermortality graphs COVID-19 mortality rates across genders.
 ```{r}
@@
 ```
-
+@fig-genderhospitalisation shows the hospitalisation, ICU, and intubation rates for different genders.
 ```{r}
@@
 ```
-
+@fig-gendertimemortality graphs monthly aggregated mortality rates across genders over time.
 ```{r}
-#| label: fig-gender-mortality-time-line
+#| label: fig-gendertimemortality
 #| fig-cap: "Mortality Rate Over Time by Gender"
-#| fig.width: 7
-#| fig.height: 4
+#| fig.width: 6.8
+#| fig.height: 3.75
 #| fig.align: center
@@
 
-Understanding the effect age has on COVID-19 is extremely important both for potential victims and those in charge of policy to put in place correct and appropriate health measures for different ages.
+Understanding the effect COVID-19 has on different age groups is extremely important both for potential victims and those in charge of policy to put in place appropriate health measures.
 
 Looking at @fig-agegroupmortality, we see a clear message - age group and mortality are positively correlated.
-More specifically, mortality past the ages of 50-59 shoots from under half a percent to 1.72% for 60-69 years of age.
+Specifically, mortality past the ages of 50-59 shoots from under half a percent to 1.72% for those 60-69 years of age.
 This only worsens as individuals get older.
-For the age groups 70-79 and 80-89, the mortality rate if one contracts COVID-19 shoots to 4.14% and 7.50% respectively.
-Moreover, those who contract COVID-19 at ages of 90 and above have a mortality rate of almost 10% (9.62%).
+For the age groups 70-79 and 80-89, the mortality rate if one contracted COVID-19 shoots to 4.14% and 7.50% respectively.
+Moreover, those who contracted COVID-19 at ages of 90 and above had a mortality rate of almost 10% (9.62%).
+In contrast, those younger than 50 at the time of contraction had mortality rates under 0.5%.
+In fact, those 19 and younger had a mortality rate of only 0.007%.
+Further, we see in @fig-fatalities-pie that those above 60 made up the vast majority of fatalities: 92.6%.
+Ages 30-59 made up 7.2% of fatalities, with less than 0.2% of fatalities coming from those younger than 30.
 
-In contrast, those younger than 50 at the time of contraction have mortality rates under 0.5%.
-In fact, those 19 and younger have a mortality rate of only 0.007%.
-Clearly, age and mortalty are very strongly correlated.
-
-Perhaps of more relevance are chances of hospitalisation, admission to the ICU, and intubation.
-In reference to @fig-agegrouphospitalisation, we see that the older population is much more susceptible to having more severe cases of COVID-19.
+Additionally, in reference to @fig-agegrouphospitalisation, we see that the older population is much more susceptible to having more severe cases of COVID-19.
 For all ages above 70, the chance of hospitalisation is above 15%, peaking at 19.46% for 80-89 year olds.
@@
 This increases with age: hospital admission rates for those with COVID-19 is 1.15% for 30-39 year olds, 2.13% for 40-49 year olds, 4.15% for 50-59 year olds, and 8.91% for 60-69 year olds.
+ICU rates and intubation rates show similar trends - as an individual gets older, COVID-19 affects them more severely.
+This data is even more harrowing when paired with the findings from @fig-treatmentoutcomes, which shows a 17.7% risk of mortality if hospitalised from COVID-19, a 39% chance of mortality if placed in the ICU, and almost a 50% chance of mortality if intubated.
 
-discussion about weaker immune systems with age (bring in some literature review)
+As individuals age, there is a gradual decline in immune function, known as immunosenescene, which impairs the body's ability to recognize and eliminate pathogens, like SARS-CoV 2, the pathogen responsible for COVID-19.
+Additionally, our immune systems suffer a 'chronic increase in systemic inflammation called inflammaging' as we grow older [@age_covid_lit].
+The article also notes age-associated comorbidities that may accelerate epigenetic age, and increase susceptibility to COVID-19.
+These could include higher rates of cardiovascular disease, diabetes, and other age-related diseases as one ages.
 
@@
 
-Similar to age, understanding differences in outcomes for genders matter greatly to virologists and as a matter of public safety.
-Looking at @fig-gendermortality, we find that males that contract COVID-19 have a much higher mortality rate when compared to females (1.47% vs 1.06%).
-According to this data, males who contract COVID-19 are almost 1.4 times more likely to die than females.
+Similar to age, understanding differences in COVID-19 outcomes for genders matter greatly to virologists and as a matter of public safety.
 
-Similarly, looking at @fig-genderhospitalisation, we see that males are more likely to be hospitalised, put in the ICU, and intubated.
+Looking at @fig-gendermortality, we find that males that contracted COVID-19 had a much higher mortality rate when compared to females (1.47% vs 1.06%).
+In other words, males who contracted COVID-19 were almost 1.4 times more likely to die than females.
+Further, @fig-genderhospitalisation tells us that males were more likely to be hospitalised, put in the ICU, and intubated - almost 5.5% of males who contract COVID-19 were hospitalised, compared to 4% for females.
+1% of males had to be placed in the ICU, with almost 0.6% having to be intubated, while females had much lower ICU rates (0.53%) and intubation rates (0.28%).
+Again, in conjunction with the results from @fig-treatmentoutcomes (which relates hospitalisation outcomes to mortality), this is an important distinction, further hammering home the point that males were more susceptible to worse outcomes if they contracted COVID-19.
 
-discussion about how Covid-19 affects the male body vs female; lit and scientific review; is this a worldwide trend?
+In looking at scientific literature, an article from the University of Toronto notes that differences in chromosomal makeup largely explains the increased severity for males with COVID-19.
+The ACE2 receptor, which SARS-CoV 2 uses to enter cells, is located on the X Chromosome.
+Females, who have XX chromosomes, have an extra copy of ACE2, effectively have more protection against inflammation and organ damage during infection.
+Males on the other hand, only have 1 copy of the ACE2 receptor, as they have XY chromosomes.
+Because of this, 'there is not enough of the protein to fulfil its usual function of tampering down inflammation and preventing organ damage' [@uoft_covid_article].
+Importantly, the study found that men infected with SARS-CoV 2 who externally supplemented ACE2 daily through an inhaler had less virus in their lungs, less lung injury, and higher estrogen signalling - protecting them from more severe COVID-19 outcomes.
+
+The significance of this discovery cannot be understated - Professor Zhang, the leader of the study imagines a world where inhaling a puff of ACE2 before entering high-risk situations is commonplace for men, to better protect them from contracting COVID-19 [@uoft_covid_article].
 
@@
 
-Reference the importance of @fig-treatmentoutcomes - those intubated have almost a 50% chance of death. (what?!)
+@fig-agetimemortality and @fig-gendertimemortality were included in this report to illustrate broader takeaways regarding the implementation of vaccinations on overall mortality rates.
+In both graphs, we find that mortality, aggregated over a month, is relatively high during the start of the pandemic in 2020.
+This is especially true for the older population - the pink, cyan, and yellow-green line graphs in @fig-agetimemortality.
+The erratic behaviour in mortality only normalises after 2022.
+In @fig-gendertimemortality, on the other hand,  male and female mortality seemed to remain close to 2.5% after normalizing in 2021, while other genders experienced greater volatility toward the start of the pandemic and similarly normalised post 2021.
+This was likely due to low observations for other genders, increasing variance in results.
+
+Interestingly, this normalisation in mortality rates to reasonable levels coincides with vaccination efforts in Toronto.
+Health Canada first approved a Pfizer-BioNTech on December 9th, 2020, which was first administered on December 14th, 2020 [@infobase_vaccines].
+From this mark on, we see a gradual decrease in mortality in both figures.
+The 50% threshold of the Canadian population received at least 1 dose of a COVID-19 vaccine was crossed on May 22, 2021, a huge milestone in vaccination efforts.
+At this point, we see very low mortality rates across genders in @fig-gendertimemortality, and a continued downward trend in mortality across age groups.
+Further, the spike in mortality created by the Omicron variant in late 2021 seen in @fig-agetimemortality quickly dies down as new booster shot programs were introduced during these months [@globalnews].
+
+The parallels between a decrease in mortality with the introduction of vaccinations show the positive effect vaccines can have on protecting the population from viruses.
+This is incredibly important for policymakers and virologists as they learn from the impact of COVID-19 to better prepare for future health crises.
 
@@
 
-Measurement weaknesses (stopped collecting data), might be important to look at selection biases (are there more male cases than females? what about Other genders? are they affect results?
+As referenced in @sec-data, the quality and depth of reporting for this data set decreased as less funding for COVID-19 resources decreased.
+This may have resulted in recent data being less complete than previous years, and cause some concerns for reliability when searching for temporal trends.
+Additionally, data may not be fully representative of the population due to under-reporting of COVID-19 cases.
+The non-mandatory nature of reporting may introduce selection biases, as cases with higher severity are more likely to be documented and reported by Toronto Public Health.
+
+An interesting next step would be to investigate comorbidity data in conjunction with this data set, to see how different comorbodities affect outcomes, and how comorbidities vary with age.
+As discussed in @sec-results, current scientific consensus agrees that the prevalence of pre-existing conditions worsens immune system responses and COVID-19 outcomes.
+Studying this theory in the context of this data might be useful to Toronto's public health systems in preparing and caring for those with comorbidities, during a public health crisis.
 
